How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?

### 2021-09-27 analysis/paper/paper.Rmd
Excluded data for SPD, adjusted raw material and artifact composition plots

After the first Paleolithic sites such as Dongkwanjin and Yondaebong were discovered during the mid-1930s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were]. We selected sites dated to before and after the transition period (42-35 ka) to analyze the process of technological change. This resulted in a sample of 33 assemblages from `r nrow(korean_archaeological_site_locations)` sites spanning 49-24 ka (Figure \@ref(fig:korea-map)). We identified multiple assemblages in a site where culturally sterile deposits separated artifact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artifacts. For example, the Hwadaeri site has three cultural layers. The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite. The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage. This layer was dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL. The uppermost layer, dated to 22,000±100 BP by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence]. We obtained the data from published excavation reports held by provincial museums.

@@

There is a debate about whether past population growth has played a role in technological change [@shennan2001demography; @kline2010population]. @shennan2009pattern observed a correlation between population decrease and technological simplification in his study of lithic arrowheads from the southern Scandinavian Mesolithic. He found that the lithics became less complex, and point shape changed as the population level dropped. On the other hand, @collard2013population argue that regional archaeological data do not support the population-driven models. For example, @buchanan2016drivers concluded that changes in point types from \>13,000 years ago to 400 years ago in Texas were related to environmental risk, and not dependent on population size. We explored SPD to test for correlations between population, environmental proxies and technological change. We used the R package 'rcarbon' [@cremaanalysing] to perform statistical analyzes and conduct model testing using Monte Carlo methods [@timpson2014reconstructing; @crema2016summed; @palmisano2020holocene]. We generated exponential, uniform, linear, and logistic models and evaluated the model goodness-of-fit by applying Akaike's Information Criterion (AIC) as a selection criterion [@sakamoto1986akaike; @bevan2017holocene; @riris2019widespread].

>>>>

After the first Paleolithic sites such as Dongkwanjin and Yondaebong were discovered during the mid-1930s, more than 200 Paleolithic sites have been discovered in South Korea [@norton2000current; @lee2017korean; @lee2019were]. We selected sites dated to before and after the transition period (42-35 ka) to analyze the process of technological change. This resulted in a sample of 33 assemblages from `r nrow(korean_archaeological_site_locations)` sites spanning 49-24 ka (Figure \@ref(fig:korea-map)). We identified multiple assemblages in a site where culturally sterile deposits separated artifact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artifacts. For example, the Hwadaeri site has three cultural layers. The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite. The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage. This layer was dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL. The uppermost layer, dated to 22,000±100 BP by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence]. We obtained the data from published excavation reports held by provincial museums.

@@

There is a debate about whether past population growth has played a role in technological change [@shennan2001demography; @hosfield2005individuals;  @kline2010population]. @shennan2009pattern observed a correlation between population decrease and technological simplification in his study of lithic arrowheads from the southern Scandinavian Mesolithic. He found that the lithics became less complex, and point shape changed as the population level dropped. By proposing the generalized null demographic model built on cultural transmission theory, @lycett2010demographic also claimed that relatively lower effective population sizes in East Asia could be strongly associated with a lower frequency of handaxe sites, a lower percentage of bifaces, and morphological difference between East Asian handaxes and classic Acheulean artifacts. They also suggested that a lack of Levallois technologies in East Asia might be similar to the previous example of handaxe. On the other hand, @collard2013population argue that regional archaeological data do not support the population-driven models. For example, @buchanan2016drivers concluded that changes in point types from \>13,000 years ago to 400 years ago in Texas were related to environmental risk, and not dependent on population size. We explored SPD to test for correlations between population, environmental proxies and technological change. We used the R package 'rcarbon' [@cremaanalysing] to perform statistical analyzes and conduct model testing using Monte Carlo methods [@timpson2014reconstructing; @crema2016summed; @palmisano2020holocene]. We generated exponential, uniform, linear, and logistic models and evaluated the model goodness-of-fit by applying Akaike's Information Criterion (AIC) as a selection criterion [@sakamoto1986akaike; @bevan2017holocene; @riris2019widespread].
